---
title: "Critical staffing shortages"
author: "Austin Fast"
date: "2/15/2022"
output: html_document
---

```{r}
#Load libraries 
library(tidyverse)
library(janitor)
library(lubridate)
```

COVID-19 Reported Patient Impact and Hospital Capacity by State Timeseries
https://healthdata.gov/Hospital/COVID-19-Reported-Patient-Impact-and-Hospital-Capa/g62h-syeh

Description of fields is at https://www.hhs.gov/sites/default/files/covid-19-faqs-hospitals-hospital-laboratory-acute-care-facility-data-reporting.pdf

Line 24 defines critical staffing shortages: Each facility should identify staffing shortages based on their facility needs and internal policies for staffing ratios. The use of temporary staff does not count as a staffing shortage if staffing ratios are met according to the facility’s needs and internal policies for staffing ratios.

Pull historical data into R
```{r}
hospital <- read_csv("https://healthdata.gov/api/views/g62h-syeh/rows.csv?accessType=DOWNLOAD")
```


This calculates by state and date how many of the hospitals report a critical staffing shortage and how many of them responded.

*Hospitals no longer have to report if they have a critical staffing shortage today, so the percent reporting figure has dropped sharply over the past month. 
*Some states (e.g., Texas) had already been doing very poorly at reporting

```{r}
hospital2 <- hospital %>% 
  group_by (date, state) %>%
  select (date, state, critical_staffing_shortage_today_yes, critical_staffing_shortage_today_no, critical_staffing_shortage_today_not_reported) %>%
  mutate (total = critical_staffing_shortage_today_yes + critical_staffing_shortage_today_no + critical_staffing_shortage_today_not_reported,
          pct_yes = critical_staffing_shortage_today_yes / (critical_staffing_shortage_today_yes + critical_staffing_shortage_today_no), 
          pct_reporting = (critical_staffing_shortage_today_yes + critical_staffing_shortage_today_no) / total)

hospital2 %>%
  filter (state == "TX") %>%
  #filter (date > as.Date("2021-12-31")) %>%
  ggplot (aes(x = date, 
              y = pct_yes,
              #fill = state
              )) +
    geom_line() 

#Check how states are doing
hospital2 %>%
  filter (date == (Sys.Date()-5)) %>%
  arrange (desc(pct_yes)) %>%
  View
```

This calculates a national rate by date of how many of the hospitals report a critical staffing shortage and how many of them responded.
*Hospitals no longer have to report if they have a critical staffing shortage today, so the percent reporting figure has dropped sharply over the past month. 
*Some states (e.g., Texas) had already been doing very poorly at reporting

```{r}
hospital3 <- hospital %>% 
  group_by (date) %>%
  select (date, state, critical_staffing_shortage_today_yes, critical_staffing_shortage_today_no, critical_staffing_shortage_today_not_reported) %>%
  summarize (total = sum(critical_staffing_shortage_today_yes) + sum(critical_staffing_shortage_today_no) + sum(critical_staffing_shortage_today_not_reported),
          pct_yes = sum(critical_staffing_shortage_today_yes) / (sum(critical_staffing_shortage_today_yes) + sum(critical_staffing_shortage_today_no)), 
          pct_reporting = (sum(critical_staffing_shortage_today_yes) + sum(critical_staffing_shortage_today_no)) / total) %>%
  #arrange (desc(date))%>%
  arrange (date) %>% 
  mutate (#days_low_vax = RcppRoll::roll_sum(dplyr::lag(low_vax),14, fill=NA, align="right"),
          #days_low_st_vax = RcppRoll::roll_sum(dplyr::lag(low_st_vax),14, fill=NA, align="right"),
          roll_avg = zoo::rollmean(pct_yes, k = 7, fill = NA, align = "right")* 100) 

hospital3 %>%
  ggplot (aes(x = date, 
              y = pct_yes,
              #fill = state
              )) +
    geom_line() 
```

This calculates a national rate of the hospitals anticipate a critical staffing shortage in a week's time and how many of them responded.

*Hospitals MUST report if they're expecting a critical staffing shortage within a week.
*Percent reporting climbs after July 2020, so this filters out early data with few hospitals reporting
```{r}
hospital4 <- hospital %>% 
  #group_by (date, state) %>% #USE THIS TO GROUP BY DATE AND STATE INSTEAD OF NATIONAL RATE
  group_by (date) %>%
  select (date, state, critical_staffing_shortage_anticipated_within_week_yes, critical_staffing_shortage_anticipated_within_week_no, critical_staffing_shortage_anticipated_within_week_not_reported) %>%
  #sum up three possible responses 
  summarize (tot_report =  sum(critical_staffing_shortage_anticipated_within_week_yes) +  sum(critical_staffing_shortage_anticipated_within_week_no),
             total = sum(critical_staffing_shortage_anticipated_within_week_yes) + sum(critical_staffing_shortage_anticipated_within_week_no) + sum(critical_staffing_shortage_anticipated_within_week_not_reported),
             #Of those that responded, calculate percent responding yes
          pct_yes = sum(critical_staffing_shortage_anticipated_within_week_yes) / (sum(critical_staffing_shortage_anticipated_within_week_yes) + sum(critical_staffing_shortage_anticipated_within_week_no)), 
             #Calculate percent that responded
          pct_reporting = (sum(critical_staffing_shortage_anticipated_within_week_yes) + sum(critical_staffing_shortage_anticipated_within_week_no)) / total) %>%
  arrange (desc(date)) %>%
  filter (date > as.Date("2020-07-31"))
```

Calculate national 7-day rolling averages of the percent anticipating a critical staffing shortage and the percent of hospitals responding 
```{r}
hospital4a <- hospital4 %>%
  arrange (date) %>% 
  mutate (#days_low_vax = RcppRoll::roll_sum(dplyr::lag(low_vax),14, fill=NA, align="right"),
          #days_low_st_vax = RcppRoll::roll_sum(dplyr::lag(low_st_vax),14, fill=NA, align="right"),
    x7_day_rolling_reporting = zoo::rollmean(tot_report, k = 7, fill = NA, align = "right"),
          x7_day_rolling_avg_pct_hospitals_anticipating_shortage = zoo::rollmean(pct_yes, k = 7, fill = NA, align = "right")* 100,
          x7_day_rolling_avg_pct_hospitals_reporting = zoo::rollmean(pct_reporting, k = 7, fill = NA, align = "right")* 100) %>%
  rename (pct_hospitals_anticipating_shortage = pct_yes,
          pct_hospitals_reporting = pct_reporting) %>%
  arrange (desc(date)) #rearrange, most recent first
```

Graph national rate of hospitals anticipating a critical staffing shortage.

Line in digital story reads, "Two years of the pandemic have pushed health care workers to the brink, with federal data showing that critical staffing shortages now plague one of every four hospitals nationwide."

Rate has fallen below 25% since writing the draft, so may need to edit depending on publish date.

```{r}
hospital4a %>%
  #filter (state == "TX") %>%
  filter (date > as.Date("2020-07-31")) %>%
  ggplot (aes(x = date)) +
    geom_line(aes(y = x7_day_rolling_avg_pct_hospitals_anticipating_shortage), color = "darkred") + 
    geom_line(aes(y = x7_day_rolling_avg_pct_hospitals_reporting), color="steelblue", linetype="twodash") +
    labs(x="Date", 
         y="% of hospitals (7-day rolling average)",
         title="Hospitals anticipating critical staffing shortages within a week", 
         #subtitle="", 
         #fill="95% CI for trendline",
         caption="Source: U.S. Department of Health & Human Services") + 
  scale_x_date(#limits = c(as.Date("2006-12-31"), as.Date("2014-09-30")), 
               breaks = seq(as.Date("2020-08-01"), as.Date("2022-03-31"), "month"),
               date_labels = "%b %Y") +
  theme(axis.text.x=element_text(angle=45, hjust=1)) +
  scale_y_continuous(limits = c(0, 30),
                     breaks = seq(0, 30, by = 5), 
                     labels = paste0(seq(0, 30, by = 5), "%")) +
  #geom_vline(xintercept = ) +
  # Shade area under y_lim
  #geom_rect(aes(xmin = as.Date("2021-09-01"), xmax = as.Date("2021-10-01"), ymin = 0, ymax = Inf), 
  annotate("rect", xmin=as.Date("2020-11-01"), xmax=as.Date("2021-02-01"), ymin=0, ymax=Inf, alpha=0.2, fill="red") +
  annotate(geom="text", x=as.Date("2020-12-15"), y=1, label="1st winter wave", size =3) +
  annotate("rect", xmin=as.Date("2021-08-01"), xmax=as.Date("2021-10-01"), ymin=0, ymax=Inf, alpha=0.2, fill="red") +
  annotate(geom="text", x=as.Date("2021-09-01"), y=1, label="Delta wave", size =3) +
  annotate("rect", xmin=as.Date("2021-12-15"), xmax=as.Date("2022-02-01"), ymin=0, ymax=Inf, alpha=0.2, fill="red") +
  annotate(geom="text", x=as.Date("2022-01-08"), y=2, label="Omicron\nwave", size =3) 
#  ylim (0, 30)

#write_csv(hospital4a, "hhs_hospitals_anticipating_critical_staffing_shortages.csv")
```

